Processing large surveys with package MRP-KIT

### 2021-07-10 case-study/Mapping_CCES_to_ACS.Rmd
checkpointing; created SurveyData objects

```{r ccesDownload, echo=FALSE, out.width = "75%", fig.cap="Harvard Dataverse download page"}
library(knitr)
include_graphics("dataverse_cces_2018_download.png")
```

@@

cces_2018_subset <- subset(cces_2018, keep_rows, keep_cols)
# `subset` drops attributes, keep attribute "var.labels"
attr(cces_2018_subset, "var.labels") <- attr(cces_2018, "var.labels")

# cleanup factor variables; drop unused category labels
for (name in names(cces_2018_subset)) {
    if (is.factor(cces_2018_subset[[name]]))  {
        cces_2018_subset[[name]] = droplevels(cces_2018_subset[[name]])
    }
}
```

@@

```{r ipumsExtract, echo=FALSE, out.width = "75%", fig.cap="IPUMS requested ACS variables"}
library(knitr)
include_graphics("ipums_acs_2018_match_cces.png")
```

@@

```{r ipumsDownload, echo=FALSE, out.width = "75%", fig.cap="IPUMS service download page"}
library(knitr)
include_graphics("ipums_download.png")
```

@@

The CCES is a survey of _national adults_, i.e., eligible
voters over the age of 18 but the ACS is a survey over the entire
US population, including non-citizens.
In order to use the ACS for poststratification, we drop all
records for individuals under the age of 18 and non-citizens.

```{r}
table(as_factor(acs_2018_subset$CITIZEN))
# drop underage, non-citizens
citizen = as_factor(acs_2018_subset$CITIZEN) != "Not a citizen"
over_18 = acs_2018_subset$AGE > 17
acs_2018_subset = acs_2018_subset[(over_18 & citizen) , ]  
```

@@

# ACS data are structured objects, ipums_val_labels returns a tibble
STATEFIP <- ipums_val_labels(acs_2018_subset$STATEFIP)[[2]][1:51]
cat(inputstate[1:5], sep=", "); cat(STATEFIP[1:5], sep=", ")
table(inputstate == STATEFIP)
```

We need to create a SurveyQuestion object, as above,  but it would be
extremely tedious to do this as a long list of tag, value pairs:
```
values_map=list("Alabama" = "Alabama", ..., "Wyoming" = "Wyoming")
```
The `values_map` is stored as an [R Pairlists](https://github.com/hadley/r-internals/blob/master/pairlists.md).
This is actually a labelled list, not a list of pairs, i.e.,
the expression `list("tag1" = "value1", "tag2" = "value2")`
creates a list of length 2 where each element is a single named item.
```{r}
foo <- list("tag1" = "value1", "tag2" = "value2")
foo
```
To create this list programmatically,
you must create a list of values and a list of labels,
then set the names of the values list to the labels.
```{r}
# create list, then set names (uppercase)
mapping <- as.list(STATEFIP)
names(mapping) <- toupper(inputstate)
mapping[1:3]
# combine the two steps into a one-liner (lowercase)
mapping = as.list(setNames(STATEFIP, tolower(inputstate)))
mapping[1:3]
```

Having established the values map, we create the corresponding SurveyQuestion object.
We use accessor methods `name()`, `col_names()` and `values()` to check our work.
```{r}
state_mapping = as.list(setNames(STATEFIP, inputstate))
q_state <- SurveyQuestion$new(name = "state",
	col_names = c("inputstate", "STATEFIP"),
	values_map= state_mapping)
q_state$name()
q_state$col_names()
head(q_state$values())
```

@@

This is borne out by the CCES, where the question for variable "race" is
"What racial or ethnic group best describes you?",
and includes "Hispanic" as a possible response.
The ACS variable "RACE" has a different set of responses:
```{r}
names(table(cces_2018_subset$race))
ipums_val_labels(acs_2018_subset$RACE)
````

@@

A further possibility is to similarly combine the CCES variables "race" and "hispan" 
so that the question of "Hispanic origin" is "Yes", the race-hispan label is "Hispanic".
Were we to do this, we would create a new variable "race2" and add it to the
CCES subset so that we could evaluate both classifications.
```{r}
# Population percentages in CCES variable "race"
prop.table(table(cces_2018_subset$race))
# Intersection (and) vs. union (or) of race == "Hispanic" and hispanic == "Yes"
length(which(cces_2018_subset$race == "Hispanic" & cces_2018_subset$hispan == "Yes"))
length(which(cces_2018_subset$race == "Hispanic" | cces_2018_subset$hispan == "Yes"))
# create column which adjusts race labels
cces_race2 <- cces_2018_subset$race
cces_race2[cces_2018_subset$hispan == "Yes"] <- "Hispanic"
prop.table(table(cces_race2))
# add to acs_2018_subset data.frame - respect column types
cces_2018_subset$race2 = cces_race2
```

@@

```{r, error=TRUE}
# OK
marstat_mapping = list(
    "Never married" = "Never married/single",
    "Domestic / civil partnership" = "Never married/single",
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)
q_marstat$name()
q_marstat$col_names()
q_marstat$values()

@@

## Measurement recorded in different units and not binned:  age

@@

@@

The size of the postratification table is the product of the number of categories per demographic.
The more fine-grained the classification, the larger the postratification table.
In this example, we have variables:
`q_sex` (2 values), `q_state` (51 values), `q_educ` (6 values), `q_race` (8 values), `q_marstat` (5 values), `q_age` (many possible values - e.g. 4 or 7).
The resultng poststratification table will have at least 97,920 cells.

@@

>>>>

```{r ccesDownload, echo=FALSE, out.width="75%", fig.cap="Harvard Dataverse download page"}
library(knitr)
include_graphics("dataverse_cces_2018_download.png")
```

@@

cces_2018_subset <- subset(cces_2018, keep_rows, keep_cols)
# `subset` drops attributes, add back the ones we need
attr(cces_2018_subset, "var.labels") <- attr(cces_2018, "var.labels")[keep_cols]

cces_2018_subset <- droplevels(cces_2018_subset)  # remove unused categories
```

@@

```{r ipumsExtract, echo=FALSE, out.width="75%", fig.cap="IPUMS requested ACS variables"}
library(knitr)
include_graphics("ipums_acs_2018_match_cces.png")
```

@@

```{r ipumsDownload, echo=FALSE, out.width="75%", fig.cap="IPUMS service download page"}
library(knitr)
include_graphics("ipums_download.png")
```

@@

The CCES is a survey of _national adults_, i.e., eligible
voters over the age of 18 but the ACS is a survey over the entire
US population, including non-citizens.
We would like to poststratify to a representative survey population,
i.e., the ACS micro-census, but that subset of the general population
who meet the CCES survey criteria, i.e., elibible voters living in
one of the 50 states or Washington DC.
Therefore,  we drop all records for individuals under the age of 18,
non-citizens, or residing outside of the 50 states or DC.

```{r}
table(as_factor(acs_2018_subset$CITIZEN))
# drop underage, non-citizens
citizen = as_factor(acs_2018_subset$CITIZEN) != "Not a citizen"
over_18 = acs_2018_subset$AGE > 17
state_or_DC = acs_2018_subset$STATEFIP < 52
acs_2018_subset = acs_2018_subset[(over_18 & citizen & state_or_DC) , ]  
```

@@

# ACS dataframe is haven_labelled, treat as factor, only keep states and DC
STATEFIP_levels <- levels(as_factor(acs_2018_subset$STATEFIP))[1:51]
cat("CCES:", inputstate[1:5], sep=", ");
cat("ACS:", STATEFIP_levels[1:5], sep=", ")
table(inputstate == STATEFIP_levels)
```

We need to create a SurveyQuestion object, as above,  but it would be
extremely tedious to do this as a long list of tag, value pairs:
```
values_map=list("Alabama" = "Alabama", ..., "Wyoming" = "Wyoming")
```
The `values_map` is stored as an [R Pairlists](https://github.com/hadley/r-internals/blob/master/pairlists.md).
This is actually a labelled list, not a list of pairs, i.e.,
the expression `list("tag1" = "value1", "tag2" = "value2")`
creates a list of length 2 where each element is a single named item.
```{r}
foo <- list("tag1" = "value1", "tag2" = "value2")
foo
```
To create this list programmatically,
you must create a list of values and a list of labels,
then set the names of the values list to the labels.
```{r}
# create list, then set names (uppercase)
mapping <- as.list(STATEFIP_levels)
names(mapping) <- toupper(inputstate)
mapping[1:3]
# combine the two steps into a one-liner (lowercase)
mapping = as.list(setNames(STATEFIP_levels, tolower(inputstate)))
mapping[1:3]
```

Having established the values map, we create the corresponding SurveyQuestion object.
We use accessor methods `name()`, `col_names()` and `values()` to check our work.
```{r}
state_mapping = as.list(setNames(STATEFIP_levels, inputstate))
q_state <- SurveyQuestion$new(name = "state",
	col_names = c("inputstate", "STATEFIP"),
	values_map= state_mapping)
q_state$name()
q_state$col_names()
head(q_state$values())
```

@@

This is borne out by the CCES, where the question for variable "race" is
"What racial or ethnic group best describes you?",
and includes "Hispanic" as a possible response.
The ACS variable "RACE" has a different set of responses:
```{r}
names(table(cces_2018_subset$race))
ipums_val_labels(acs_2018_subset$RACE)
```

@@

A further possibility is to similarly combine the CCES variables "race" and "hispan" 
so that the question of "Hispanic origin" is "Yes", the race-hispan label is "Hispanic".
Were we to do this, we would create a new variable "race2" and add it to the
CCES subset so that we could evaluate both classifications.
```{r}
# Population percentages in CCES variable "race"
prop.table(table(cces_2018_subset$race))
# Intersection (and) vs. union (or) of race == "Hispanic" and hispanic == "Yes"
length(which(cces_2018_subset$race == "Hispanic" & cces_2018_subset$hispan == "Yes"))
length(which(cces_2018_subset$race == "Hispanic" | cces_2018_subset$hispan == "Yes"))
# create column which adjusts race labels, add to dataframe
cces_race2 <- cces_2018_subset$race
cces_race2[cces_2018_subset$hispan == "Yes"] <- "Hispanic"
prop.table(table(cces_race2))
cces_2018_subset$race2 = cces_race2
attr(cces_2018_subset, "var.labels")[['race2']] <- "Race or Hispanic"
```

@@

```{r error=TRUE}
# OK
marstat_mapping = list(
    "Never married" = "Never married/single",
    "Domestic / civil partnership" = "Never married/single",
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)
q_marstat$name()
q_marstat$col_names()
q_marstat$values()

@@

## Measurement recorded in different units and not binned:  brithyear and age

@@

```{r}
# use R `cut` functon, choose breakpoints, labels
cohort_breaks = c(1900, 1935, 1945, 1955, 1965, 1975, 1985, 2000)
cohort_labels=c("before-1936",
         "1936-1945",
         "1946-1955",
         "1956-1965",
         "1966-1975",
         "1976-1985",
         "after 1986")

# add to CCES
cces_2018_subset$birthyr_cohort = cut(cces_2018_subset$birthyr, cohort_breaks, cohort_labels)
attr(cces_2018_subset, "var.labels")[['birthyr_cohort']] <- "Birthyear cohort"

# transform ACS age to birthyr, then add to ACS
acs_2018_subset$birthyr = 2018 - acs_2018_subset$AGE
acs_2018_subset$birthyr_cohort = to_labelled(cut(acs_2018_subset$birthyr, cohort_breaks, cohort_labels))
```

This is a straightforward 1:1 mapping
```{r}
birthyr_cohort_mapping = as.list(setNames(cohort_labels, cohort_labels))
q_birthyr <- SurveyQuestion$new(name = "birthyear",
	col_names = c("birthyr_cohort", "birthyr_cohort"),
	values_map= birthyr_cohort_mapping)
q_birthyr$name()
q_birthyr$col_names()
head(q_birthyr$values())
```

@@

The size of the postratification table is the product of the number of categories per demographic.
The more fine-grained the classification, the larger the postratification table.
In this example, we have variables:
`q_sex` (2 values), `q_state` (51 values), `q_educ` (6 values), `q_race` (8 values), `q_marstat` (5 values), `q_birthyr` (many possible values - e.g. 4 or 7).
The resultng poststratification table will have at least 97,920 cells.

@@

A SurveyData object requires a dataframe, a question list, and a response list.
A question list is a list of tag-value pairs where the tag (name) is the dataframe column name and the value is a description of corresponding survey question.
For the CCES, the dataframe column descriptions are lists in the attrbute "var.labels".
As before, we create this list programatically using the `as.list(setNames(tags,values))` trick.
```{r}
cces_2018_questions <- as.list(setNames(attr(cces_2018_subset,"var.labels"), names(cces_2018_subset)))
head(cces_2018_questions)
```
For the ACS, we use the function `ipumsr::ipums_var_label` to get a short description of the variable.
```{r}
acs_2018_var_labels <- c()
for (name in names(acs_2018_subset)) {
    acs_2018_var_labels <- c(acs_2018_var_labels, ipums_var_label(acs_2018_subset[[name]]))
}
acs_2018_questions <- as.list(setNames(acs_2018_var_labels, names(acs_2018_subset)))
head(acs_2018_questions)
```

A response list is a list of name-value pairs which map the column name to the set of
possible response values.
We can create this list using a for loop, or by using R's `apply` function.
```{r}
# create response list using a for loop
cces_2018_responses <- list()
for (name in names(cces_2018_subset)) {
    cces_2018_responses[[name]] <- levels(as.factor(cces_2018_subset[[name]]))
}
head(cces_2018_responses)

acs_2018_responses <- list()
for (name in names(acs_2018_subset)) {
    acs_2018_responses[[name]] <- levels(as_factor(acs_2018_subset[[name]]))
}
acs_2018_responses[["MARST"]]
acs_2018_responses[["race2"]]
acs_2018_responses[["birthyr_cohort"]]

# cces_2018_subset is a dataframe; R's `apply` function would also work
# cces_2018_responses <- apply(cces_2018_subset, 2, function(x) levels(as.factor(x)))
# if we coerce acs_2018_subset to dataframe, can do the same
# acs_2018_responses <- apply(as.data.frame(cces_2018_subset), 2, function(x) levels(as.factor(x)))
```

We use these lists to create SurveyData objects for both the CCES and ACS.
```{r}
cces_2018_data <- SurveyData$new(cces_2018_subset, cces_2018_questions, cces_2018_responses)
cces_2018_data
acs_2018_data <- SurveyData$new(acs_2018_subset, acs_2018_questions, acs_2018_responses)
acs_2018_data
```
